Rising from the Ashes II: The Bar-driven Abundance Bimodality in the Milky Way

### 2024-10-11 paper/ms.tex
Some small edits and title

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{newtxtext,newtxmath}
\usepackage{hyperref}
\usepackage{gensymb}
\usepackage{enumitem}

@@

\title{Rising from the Ashes II: Can Bar Formation Lead to Abundance Bimodalities?}

@@

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation. Following a minor post-processing adjustment -— motivated by recent findings on the star formation efficiency in TNG -- which increases the \alphaFe{} values of older star particles, we demonstrate that this galaxy exhibits a pronounced bimodal distribution in the \alphaFe{}-\FeH{} plane. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by AGN-induced quenching triggered by the formation of its bar. We predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

\begin{table}[htb]
  \centering
  \begin{tabular}{ll}
  \multicolumn{2}{c}{dark matter}                                                                  \\ \hline
  \multicolumn{1}{l|}{$M_{200}$}                                      & $4.97\times10^{12}\,\Msun$ \\
  \multicolumn{1}{l|}{$R_{200}$}                                      & $532\,\kpc$                \\
                                                                      &                            \\
  \multicolumn{2}{c}{baryons}                                                                      \\ \hline
  \multicolumn{1}{l|}{$\rhalf$}                                       & $2\,\kpc$                  \\
  \multicolumn{1}{l|}{$M_{\textrm{star}}(r<4\rhalf)$}                 & $7.6\times10^{12}\,\Msun$  \\
  \multicolumn{1}{l|}{$M_{\textrm{gas}}(r<4\rhalf)$}                  & $2.6\times10^{11}\,\Msun$  \\
  \multicolumn{1}{l|}{$\textrm{SFR}$}                                 & $0.56\,\Msunyr$            \\
  \multicolumn{1}{l|}{$M_{\textrm{BH}}$}                              & $2.9\times10^{8}\,\Msun$   \\
  \multicolumn{1}{l|}{$\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{Edd}}$} & $4.5\times10^{-5}$        
  \end{tabular}
  \caption{Summary statistics of our subhalo of interest at $z=0$ (392276). $M_{200}$ and $R_{200}$ are relative to the mean density of the universe, the SFR is for all gas bound to the subhalo, and the BH refers to the central BH.}
  \label{tab:summ}
\end{table}

@@

The quantity of particular importance to our results is the star formation efficiency (SFE) timescale $\tau_\star$:
\begin{equation}
\tau_\star \equiv \frac{M_{\textrm{gas}}}{\SFR},
\end{equation}
where $M_{\textrm{gas}}$ is the mass of the gas reservoir. This timescale can be equivalently expressed as the ratio of the corresponding surface densities $\Sigma_{\textrm{gas}}$ and $\dot{\Sigma}_{\textrm{star}}$. $\tau_\star$ is the inverse of the SFE itself, describing the rate of star formation per unit gas mass. In principle, $\tau_\star$ should vary with $M_{\textrm{gas}}$ because the observed relation between $\dot{\Sigma}_{\textrm{star}}$ and $\Sigma_{\textrm{gas}}$ is non-linear; classically, $\dot{\Sigma}_{\textrm{star}} \propto \Sigma_{\textrm{gas}}^N$ where $N \approx 1.5$ (e.g. \citealt{Schmidt1959, Kennicutt1998}). For simplicity, we hold $\tau_\star$ constant in this paper, which corresponds to a linear relation. In Section~\ref{ssec:onezone} below, we present models using $\tau_\star = 0.5$, $1$, and $5\,\Gyr$.

@@

In Figure~\ref{fig:fig1}, we showed that a time-linear $\alpha$-enhancement of older stars (formed before $z=1.5$) leads to the emergence of a pronounced chemical bimodality. This $\alpha$-enhancement corresponds to a more rapid decline in \alphaFe{} over time at high redshifts. Figure~\ref{fig:vice} suggests that a low SFE at high densities may plausibly explain the absence of a steep decline in \alphaFe{} in the fiducial TNG model.

@@

This analysis illustrates that a higher SFE at early times (high redshift) leads to a faster decrease in \MgFe{}. Recent work has suggested that the SFE in such dense regions in TNG may indeed be too low, as discussed in Section~\ref{ssec:sfe}.

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{newtxtext,newtxmath}
\usepackage{hyperref}
\usepackage{gensymb}
\usepackage{enumitem}
\usepackage{booktabs}

@@

\title{Rising from the Ashes II: The Bar-driven Abundance Bimodality in the Milky Way}

@@

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation.

    Following a minor post-processing adjustment -— motivated by recent findings on the star formation efficiency in TNG -- which increases the \alphaFe{} values of older star particles, we demonstrate that this galaxy exhibits a pronounced bimodal distribution in the \alphaFe{}-\FeH{} plane.

    This sequence results in a pronounced abundance bimodality after a post-processing adjustment of the \alphaFe{} of old stars which is designed to mimic a higher star formation efficiency in dense gas.

    The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by AGN-induced quenching triggered by the formation of its bar. We predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

% Please add the following required packages to your document preamble:
\begin{table}[]
  \centering
  \begin{tabular}{@{}ll@{}}
  \toprule
  \multicolumn{2}{c}{dark matter}                                                  \\ \midrule
  \multicolumn{1}{l|}{$M_{200}$}                      & $4.97\times10^{12}\,\Msun$ \\
  \multicolumn{1}{l|}{$R_{200}$}                      & $532\,\kpc$                \\
                                                      &                            \\
  \multicolumn{2}{c}{baryons}                                                      \\ \midrule
  \multicolumn{1}{l|}{$\rhalf$}                       & $2\,\kpc$                  \\
  \multicolumn{1}{l|}{$M_{\textrm{star}}(r<4\rhalf)$} & $7.6\times10^{12}\,\Msun$  \\
  \multicolumn{1}{l|}{$M_{\textrm{gas}}(r<4\rhalf)$}  & $2.6\times10^{11}\,\Msun$  \\
  \multicolumn{1}{l|}{$\textrm{SFR}$}                 & $0.56\,\Msunyr$            \\
  \multicolumn{1}{l|}{$M_{\textrm{BH}}$}              & $2.9\times10^{8}\,\Msun$   \\
  $\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{Edd}}$      & $4.5\times10^{-5}$         \\ \bottomrule
  \end{tabular}
  \caption{Summary statistics of our subhalo of interest at $z=0$. $M_{200}$ and $R_{200}$ are relative to the mean density of the universe, the SFR is for all gas bound to the subhalo, and the BH refers to the central BH.}
  \label{tab:summ}
  \end{table}

@@

The quantity of particular importance to our results is the star formation efficiency (SFE) timescale $\tau_\star$ (also known as the depletion time):
\begin{equation}
\tau_\star \equiv \frac{M_{\textrm{gas}}}{\SFR},
\end{equation}
where $M_{\textrm{gas}}$ is the mass of the gas reservoir. This timescale can be equivalently expressed as the ratio of the corresponding surface densities $\Sigma_{\textrm{gas}}$ and $\dot{\Sigma}_{\textrm{star}}$. $\tau_\star$ is the inverse of the SFE itself, describing the rate of star formation per unit gas mass. In principle, $\tau_\star$ should vary with $M_{\textrm{gas}}$ because the observed relation between $\dot{\Sigma}_{\textrm{star}}$ and $\Sigma_{\textrm{gas}}$ is non-linear; classically, $\dot{\Sigma}_{\textrm{star}} \propto \Sigma_{\textrm{gas}}^N$ where $N \approx 1.5$ (e.g. \citealt{Schmidt1959, Kennicutt1998}). For simplicity, we hold $\tau_\star$ constant in this paper, which corresponds to a linear relation. In Section~\ref{ssec:onezone} below, we present models using $\tau_\star = 0.5$, $1$, and $5\,\Gyr$.

@@

In Figure~\ref{fig:fig1}, we showed that a time-linear $\alpha$-enhancement of older stars (formed before $z=1.5$) leads to the emergence of a pronounced chemical bimodality. This $\alpha$-enhancement corresponds to a more rapid decline in \alphaFe{} over time at high redshifts. 
We demonstrate the physical motivation behind this $\alpha$-enhnacement in Figure~\ref{fig:vice}.

@@

This analysis illustrates that a higher SFE at early times (high-$z$) leads to a faster decline in \MgFe{}. Recent work has suggested that the SFE in such dense regions in TNG may indeed be too low, as discussed in Section~\ref{ssec:sfe}. The post-processing $\alpha$-enhancement in Section~\ref{ssec:tng} is meant to mimic a SFE correction of high-$z$ dense gas.
